---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
Load activity.csv and format the Date correctly using as.Date
```{r}
library(plyr)
data<-read.csv("activity.csv")
data$date<-as.Date(data$date,"%Y-%m-%d")                
```

## What is mean total number of steps taken per day?
Here is a histogram on the number of steps taken per pay
```{r}
        datasum<-ddply(data,.(date),summarize, steps= sum(steps))
        hist(datasum$steps, main="Number of Steps per Day",xlab="Number of Steps",breaks=15,col="red")
```

Turn data into Steps per day and then calculate the mean and median

```{r}
        ##prints mean steps per day
        mean(datasum$steps,na.rm=TRUE)
        
        ##prints median steps per day
        median(datasum$steps,na.rm=TRUE)
```


## What is the average daily activity pattern?
This histogram shows the anverage number of steps per the 5-minute interval times
```{r}
        dataave<-ddply(data,.(interval),summarize, steps=mean(steps,na.rm=TRUE))
        plot1<-plot(x=dataave$interval, y=dataave$steps,type="l",xlab="Interval",ylab="Average
                    Steps",main="Daily Average Number of Steps per  Interval",col="red")
```

So which interval has the most steps of average?

        
```{r}
        #highest value for averagesteps per day
        print(max(dataave$steps,na.rm=TRUE))
        #the interval
        dataave$interval[dataave$steps==max(dataave$steps,na.rm=TRUE)]
```

## Imputing missing values
Lets take a look at the missing data in the dataset:
```{r}
        NAs <-sapply(data, function(data) sum(length(which(is.na(data)))))
        
        print(NAs)## prints Number of NAs in each column (finding our NAs)
        
```
I used the mean of steps over each interval and day of the week to create the imputed NA values from the data set.
```{r}
         ##add col to show it is imputed or not
        dataref<-mutate(data,isImputed = is.na(data$steps))
        
        ##make a key of average steps per interval and day of the week to use to impute NA values
        dataref<-ddply(dataref[dataref$isImputed==FALSE,],.(interval,date),summarize, steps=mean(steps,na.omit=TRUE),date=weekdays(date))
        dataref<-aggregate(dataref$steps,list(dataref$interval, dataref$date),mean)
        dataref<-setNames(dataref,c("interval","date","steps"))
        
        ##create data with NAs imputed using dataref
        dataimp<-data
        dataimp$date<-weekdays(dataimp$date)
        dataimp<-mutate(dataimp,isImputed = is.na(data$steps))
        library(data.table)
        setDT(dataimp)
        setDT(dataref)
        ##Use int for steps (rounds the average to int)
        dataref$steps<-as.integer(dataref$steps)
        
        setkey(dataimp,interval,date); setkey(dataref,interval,date)
        dataimp<-
                dataimp[dataref, steps := ifelse(is.na(steps), i.steps, steps) ]
        
        dataimp<-data.frame(dataimp)
```
The NAs are now filled in.  Lets take a look
```{r}
        #find NAs in imputed data
        NAis <-sapply(dataimp, function(dataimp) sum(length(which(is.na(dataimp)))))
        print(NAis )## prints Number of NAs in dataframe (should have none)
```
Lets check out the Number of steps per day with the imputed data
```{r}
 #turn date back to normal date
        dataimp$date<-data$date
        datasumi<-ddply(dataimp,.(date),summarize, steps= sum(steps))
        hist(datasumi$steps, main="Number of Steps per Day",xlab="Number of                        
             Steps",breaks=30,col="red")
```

You can see there is much higher amount of days with 0 steps in the imputed data. The data of course now with added step values have increased toward a greater number of steps, but other than that the data maintains a similar pattern as does the not imputed data.  
Calculate the mean and median

```{r}
        ##prints mean steps per day
        mean(datasumi$steps,na.rm=TRUE)
        
        ##prints median steps per day
        median(datasumi$steps,na.rm=TRUE)
```

The two affects mentioned previously explain the shifting of the median (all the added zero values) and the average to the right (adding all the extra steps.)


## Are there differences in activity patterns between weekdays and weekends?
The weekend and weekday groups follow similar patterns when looking at the average steps per interval.  
```{r}
   
        ##seperate weekends from weekdays
        library(chron)
        library(ggplot2)
        dataimp$date<-is.weekend(dataimp$date)
        datawk<-ddply(dataimp, .(interval, date, isImputed),summarize, steps=mean(steps))
        datawk<-ddply(datawk,.(interval,date,isImputed),summarize,steps=sum(steps),
                        date=if(date==TRUE){date<-"Weekend"}else{date<-"Weekday"})
        plot.final<-qplot(x=interval,y=steps,data=datawk,geom=c("point","line"),
                        facets = date~.,xlab="Interval",ylab="Ave Steps",col="red",type="l",                           main="Average steps per interval")
        print(plot.final)
             
```
The following graph also shows the imputed values are fit in nicely with the patterns seen in the actual data.
```{r}
   
        #add col for Imputed Data
        plot.finali<-qplot(x=interval,y=steps,data=datawk,geom=c("point","line"),
                        facets = date~.,xlab="Interval",ylab="Ave Steps",type="l", col=isImputed,                           main="Average steps per interval")
        print(plot.finali)
             
```        



